---
title: "`r cross`"
output: 
    rmdformats::html_clean:
        use_bookdown: TRUE
        lightbox: TRUE
        gallery: TRUE
---

This is a report on TRD analyses of cross `r cross`.

````{r}
# get the data

library(tidytable)
library(ggplot2)
library(naturalsort)
library(scales)
library(stringr)

AF_across_genome<-fread(paste0("~/data/TRD/results/shiny/", cross, "-AF.csv.gz"))

TRD_regions<-fread(paste0("~/data/TRD/results/shiny/", cross, "-TRD_regions.csv.gz"))



````

# The two parents

````{r}

# load the Excel crosses file

crosses_xlsx=readxl::read_xlsx("~/data/trd/Crosses.xlsx", sheet=2)
cc=data.frame("Cross ID"=c(paste0("ChrisC",1:8)),
                                  "Short name 1"=c("ACP","BAP","CCD","ATE","ACK","AKE","BAH","ANG"),
                                  "Short name 2"=c("BFP","CMP","CPG","SACE_YCR","ACV","BAH","CGD","CEI"), stringsAsFactors=FALSE)
colnames(cc)=str_replace_all(colnames(cc), fixed("."), " ")
crosses_all <- bind_rows(crosses_xlsx[,c("Cross ID","Short name 1","Short name 2")],
                        cc)
                        
crosses_all_slice<-filter(crosses_all, `Cross ID` == cross)
strain_1<-pull(crosses_all_slice, `Short name 1`)
strain_2<-pull(crosses_all_slice, `Short name 2`)

count_segregants<-NA

if(cross %in% pull(crosses_xlsx, `Cross ID`)){
    count_segregants<-filter(crosses_xlsx, `Cross ID` == cross)%>%
                    pull(`Colonies in pool`)
}

# load the Victor clusters

df_Strains<-fread("~/TRD/Shiny/data/Victor/operationalTable_Full2543Sace_Clades.csv")

cluster_1<-filter(df_Strains, StandardizedName == strain_1)%>%pull(Clade)
cluster_2<-filter(df_Strains, StandardizedName == strain_2)%>%pull(Clade)

# load the ASD

ASD_comparisons<-fread("~/data/TRD/comparisons.csv")
ASD_value<-filter(ASD_comparisons, (row==strain_1 & col == strain_2) |
                                    (row==strain_2 & col==strain_1))%>%
                                    pull(dist)
ASD_value<-unique(ASD_value)

ASD_translated_value<-ASD_value*(1554384/12000000)
ASD_translated_value<-round(ASD_translated_value*100,2)

````

Cross `r cross` was done between `r strain_1` (`r cluster_1`) and `r strain_2` (`r cluster_2`). I managed to pool `r count_segregants`. 

The parental strains had an allele sharing distance of `r ASD_value`, which I think should translate into `r paste0(ASD_translated_value,"%")` distance between the genomes.


# TRD across genome

````{r}

# code below from TRD/02_TRD/01_getTRDs-GATK.R

chrs <- summarise(group_by(AF_across_genome, chr), maxpos = max(pos))
chrs <- chrs[naturalorder(chrs$chr), ]
chrs$global_pos <- cumsum(chrs$maxpos)

p<-ggplot(AF_across_genome, aes(global_pos, AD_A1 / sumCount)) +
  geom_point(alpha = 0.1, color = "grey") +
  geom_line(mapping = aes(global_pos, smoothed, color = abs(0.5 - smoothed)), inherit.aes = FALSE, linewidth = 2) +
  scale_color_viridis_c(option = "A", limits = c(0, 0.5)) +
  ylim(c(0, 1)) +
  geom_hline(yintercept = 0.5) +
  geom_vline(xintercept = chrs$global_pos) +
  geom_vline(xintercept=c(pull(TRD_regions,global_start),pull(TRD_regions,global_end)), color="red", linetype=2, alpha=0.5)+
  theme_bw(16) +
  ylab("Allele Frequency") +
  xlab("POS") +
  theme(legend.position = "none") +
  # geom_hline(yintercept = c(0.4,0.6))+
  ggtitle(cross) +
  labs(alpha = "Coverage") +
  scale_x_continuous(labels = comma)
  
TRD_regions<-mutate(TRD_regions,
                    length90=NA,
                    length80=NA,
                    length70=NA,
                    length60=NA,
                    maxStrength=NA)
                    
for(i in 1:nrow(TRD_regions)){
    TRD_regions_slice<-slice(TRD_regions, i)
    AF_across_genome_here<-filter(AF_across_genome, chr==pull(TRD_regions_slice, chr_start) & global_pos >= pull(TRD_regions_slice, global_start) & global_pos <= pull(TRD_regions_slice, global_end))
    
    TRD_regions<-mutate(TRD_regions,
                        length90=ifelse(ID==pull(TRD_regions_slice,ID),sum((pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))>=0.9 | 
                        (pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))<= 0.1),length90),
                        length80=ifelse(ID==pull(TRD_regions_slice,ID),sum((pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))>=0.8 | 
                        (pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))<= 0.2),length80),
                        length70=ifelse(ID==pull(TRD_regions_slice,ID),sum((pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))>=0.7 | 
                        (pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))<= 0.3),length70),
                        length60=ifelse(ID==pull(TRD_regions_slice,ID),sum((pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))>=0.6 | 
                        (pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))<= 0.4),length60),
              maxStrength=ifelse(ID==pull(TRD_regions_slice,ID),ifelse(max(pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))>max(pull(AF_across_genome_here,AD_A2)/pull(AF_across_genome_here,sumCount)),
                        max((pull(AF_across_genome_here,AD_A1)/pull(AF_across_genome_here,sumCount))),max((pull(AF_across_genome_here,AD_A2)/pull(AF_across_genome_here,sumCount)))),maxStrength))
}
  
````

````{r}

p

````

## List of TRD regions

The order is the same as highlighted in the plot above. IDs don't necessarily increase continuously.

````{r} 

library(reactable)

reactable(select(TRD_regions, ID, lengthSNPs, length90,
                    length80,
                    length70,
                    length60, lengthBp, 
                    maxStrength, chr_start, chr_end, global_start, global_end))

````

# Structural variants

````{r}

sv_plots_files<-paste0("~/data/trd/SV_analysis/",cross,".",c("sv_victor","mash1","mash2"),".RDS")

get_plots_if_they_exist<-function(x){
    if(file.exists(x)){
        return(readRDS(x))
    }else{
        return(NA)
    }
}

sv_plots<-lapply(sv_plots_files, get_plots_if_they_exist)
names(sv_plots)<-c("sv_victor","mash1","mash2")



````

## Victor's called SVs between parental strains

It is worth noting here that the sizes of SVs can be quite different between the parental strains and what has been merged on a population-level and is then shown here. Later, I will show quickly called SVs just between the two respective assemblies that one can then compare visually.

"TRD" regions added as well.

````{r, fig.height=8, eval=!is.na(sv_plots[["sv_victor"]])} 
sv_plots[["sv_victor"]]+theme_bw(12)
````


### Summary per TRD of % overlaps

````{r, eval=file.exists(paste0("/home/jnrunge/data/trd/SV_analysis/",cross,".SV_data.RDS"))}

SV_data<-readRDS(paste0("/home/jnrunge/data/trd/SV_analysis/",cross,
                    ".SV_data.RDS"))
                    
for(ci in unique(paste0(pull(SV_data, cross), "_",
                     pull(SV_data, ID), "_",
                       pull(SV_data, sample)))){
   SV_data_sub<-filter(SV_data, paste0(cross,"_",ID,"_",sample)==ci)
   
   SV_data_sub_TYPE_source<-select(SV_data_sub, TYPE, source)%>%distinct()
   
   missing_TYPE_source<-anti_join(All_TYPE_source_combinations, SV_data_sub_TYPE_source)
   
   new_rows<-mutate(missing_TYPE_source, cross=strsplit(ci,"_",fixed=TRUE)[[1]][1],
                   ID=strsplit(ci,"_",fixed=TRUE)[[1]][2],
                    sample=strsplit(ci,"_",fixed=TRUE)[[1]][3], LEN_rel=0)
   
   SV_data<-bind_rows(SV_data, new_rows)
      
}

SV_data_summary<-SV_data%>%filter(!is.na(TYPE))%>%
group_by(cross,ID,TYPE,source, sample)%>%summarize(sum_LEN_rel=sum(LEN_rel))
                    
reactable(SV_data_summary%>%group_by(cross,ID,TYPE)%>%summarize(LEN_rel_diff_abs=abs(sum_LEN_rel[1]-sum_LEN_rel[2]))
%>%arrange(-LEN_rel_diff_abs))

## Dotplots

To help us understand some SVs better, I also produced dotplots, from both parents perspectives.

````{r, fig.height=8, eval=!is.na(sv_plots[["mash1"]])} 
sv_plots[["mash1"]]+theme_bw(12)
````

````{r, fig.height=8, eval=!is.na(sv_plots[["mash2"]])} 
sv_plots[["mash2"]]+theme_bw(12)
````




````

# Local phylogeny

- Summary

````{r}

LP_data<-fread("/home/jnrunge/data/trd/local_phylogenies_trd_analysis/TRD_regions_with_LP_data.csv.gz")

cross_val<-cross

LP_data<-filter(LP_data,cross==cross_val)%>%select(ID,PCA_eucldist_quantile_1,PCA_eucldist_sd_multiplier_1,IBS_eucldist_quantile_1,IBS_eucldist_sd_multiplier_1,tree_changes_quantile,tree_changes_sd_multiplier,PCA_eucldist_quantile_2,PCA_eucldist_sd_multiplier_2,IBS_eucldist_quantile_2,IBS_eucldist_sd_multiplier_2)

LP_data<-pivot_longer(LP_data, -ID)

LP_data<-mutate(LP_data, value=round(value,2), abs=round(abs(value),2))

reactable(LP_data%>%arrange(-abs))

````

- Plots

## PCA / IBS-MDS

First, we can visualize the PCA and IBS-MDS of the TRD region across the 2.5K strains, highlighting the two crossed strains, blue for strain 1 and red for strain 2 (referring to up and down in the TRD plot).

````{r, results="asis"}
for(id in pull(TRD_regions, ID)){
    if(!file.exists(paste0("~/data/trd/local_phylogenies_trd_analysis/",
                cross,".",id,".","PCA",".RDS"))){
                next
                }
    cat("### TRD region ",id, "  ")
    cat("\n") 
print(readRDS(paste0("~/data/trd/local_phylogenies_trd_analysis/",
                cross,".",id,".","PCA",".RDS")))
              cat("\n")   
print(readRDS(paste0("~/data/trd/local_phylogenies_trd_analysis/",
                cross,".",id,".","IBS MDS",".RDS")))
                cat("\n\n")   
}


````

## Distance plots

Next, we want to quantify how much the TRD regions are more or less distant to other strains in the PCA / IBS-MDS above compared to the rest of the strains's genomes.

The red lines indicate the distance to the midpoint in the PCA or IBS-MDS of the TRD region, i.e. how different the strain's region is compared to the average in the 2.5K dataset. To put it into perspective, 10,000 random 50 kB regions were drawn and the focal strain's distance to the mean was taken for those as well. These random regions form the histogram in the background. The blue line shows a normal distribution of the same mean and sd as the random values. Hence, a strongly shifted red lined (shifted from median in the histogram and/or mean in the blue line) would indicate a particularly average (low value) or particularly distance (high value) TRD region.

````{r, results="asis"}
for(id in pull(TRD_regions, ID)){
    if(!file.exists(paste0("~/data/trd/local_phylogenies_trd_analysis/",
                cross,".",id,".",strain_1,".","PCA",".RDS"))){
                next
                }
    cat("### TRD region ",id, " in strain 1, ", strain_1, "  ")
    cat("\n") 
print(readRDS(paste0("~/data/trd/local_phylogenies_trd_analysis/",
                cross,".",id,".",strain_1,".","PCA",".RDS")))
              cat("\n")   
print(readRDS(paste0("~/data/trd/local_phylogenies_trd_analysis/",
                cross,".",id,".",strain_1,".","IBS",".RDS")))
                cat("\n\n")   
                
                cat("### TRD region ",id, " in strain 2, ", strain_2, "  ")
    cat("\n") 
print(readRDS(paste0("~/data/trd/local_phylogenies_trd_analysis/",
                cross,".",id,".",strain_2,".","PCA",".RDS")))
              cat("\n")   
print(readRDS(paste0("~/data/trd/local_phylogenies_trd_analysis/",
                cross,".",id,".",strain_2,".","IBS",".RDS")))
                cat("\n\n")   
}


````

## Changes in the local phylogeny

- needs to be re-analyzed as basically all TRD regions were low outliers, implying that the 50 kB region size of the random regions determined more changes than the often larger TRD regions.


# Population genomics

## Nucleotide diversity

````{r, results="asis"}
for(id in pull(TRD_regions, ID)){
    if(!file.exists(paste0("/home/jnrunge/data/trd/pop_genomics/",cross,".",id,"-relative-measures.RDS"))){
    next
    }
    rel_measures<-readRDS(paste0("/home/jnrunge/data/trd/pop_genomics/",cross,".",id,"-relative-measures.RDS"))
    cat("### TRD region ",id)
    cat("\n") 
    cat(paste0("The distorter-like strains have a pi of ", rel_measures$pi_distorter$pi_pop, " in the TRD region. That is the ", 
    rel_measures$pi_distorter$ecdf_value, " percentile, or a move of ", rel_measures$pi_distorter$standardized_measure, " SD from the mean, compared to the average of the distorters-like strains genome-wide."))
    cat("\n")
    cat(paste0("The other strains have a pi of ", rel_measures$pi_other$pi_pop, " in the TRD region. That is the ", 
    rel_measures$pi_other$ecdf_value, " percentile, or a move of ", rel_measures$pi_other$standardized_measure, " SD from the mean, compared to the average of the distorters-like strains genome-wide."))
    cat("\n")
    print(readRDS(paste0("/home/jnrunge/data/trd/pop_genomics/",cross,".",id,"-pi-plot.RDS")))
    cat("\n")
    cat("\n")
     
}


````

## Fst

````{r, results="asis"}
for(id in pull(TRD_regions, ID)){
if(!file.exists(paste0("/home/jnrunge/data/trd/pop_genomics/",cross,".",id,"-relative-measures.RDS"))){
    next
    }
    rel_measures<-readRDS(paste0("/home/jnrunge/data/trd/pop_genomics/",cross,".",id,"-relative-measures.RDS"))
    cat("### TRD region ",id)
    cat("\n") 
    cat(paste0("The population differentiation of distorter-like and other strains at the TRD locus is ", rel_measures$fst$sum_value, ". That is the ", 
    rel_measures$fst$ecdf_value, " percentile, or a move of ", rel_measures$fst$standardized_measure, " SD from the mean, compared to the average genome-wide."))
    cat("\n")
    print(readRDS(paste0("/home/jnrunge/data/trd/pop_genomics/",cross,".",id,"-Fst-plot.RDS")))
    cat("\n")
     cat("\n")
}


````

## Dxy

````{r, results="asis"}
for(id in pull(TRD_regions, ID)){
if(!file.exists(paste0("/home/jnrunge/data/trd/pop_genomics/",cross,".",id,"-relative-measures.RDS"))){
    next
    }
    rel_measures<-readRDS(paste0("/home/jnrunge/data/trd/pop_genomics/",cross,".",id,"-relative-measures.RDS"))
    cat("### TRD region ",id)
    cat("\n") 
    cat(paste0("The nucleotide divergence of distorter-like and other strains at the TRD locus is ", rel_measures$dxy$sum_value, ". That is the ", 
    rel_measures$dxy$ecdf_value, " percentile, or a move of ", rel_measures$dxy$standardized_measure, " SD from the mean, compared to the average genome-wide."))
    cat("\n")
    print(readRDS(paste0("/home/jnrunge/data/trd/pop_genomics/",cross,".",id,"-Dxy-plot.RDS")))
    cat("\n")
     cat("\n")
}


````


## Tajima's D

To understand whether the TRD region is showing signals of selection across all strains, I also calculated Tajima's D. *Note that this is not on a subset of e.g. distorter-like strains, but across all strains of the 2.5K matrix.*

````{r, results="asis"}
TD_values<-fread("/home/jnrunge/data/trd/TD_data/TD_plot_data.csv.gz")
cross_value<-cross
for(id in pull(TRD_regions, ID)){
    TD_values_slice<-filter(TD_values, cross==cross_value, ID==id)
    cat("### TRD region ",id)
    cat("\n") 
    cat(paste0("The TD value in the TRD region across all strains is ", pull(TD_values_slice, TD), ", which represents a percentile of ",
    pull(TD_values_slice, TD_ecdf), " and an SD shift of ", pull(TD_values_slice, TD_sd_multi)))
     cat("\n") 
    print(readRDS(paste0("/home/jnrunge/data/trd/TD_data","/",cross,".", id,".",
                     "TD_plot",".RDS")))
     cat("\n") 
      cat("\n") 

}
````

## Linkage disequillibrium

Another important hint could be linkage disequillibrium, which I have also calculated across all strains. *Note that this is not on a subset of e.g. distorter-like strains, but across all strains of the 2.5K matrix.*

````{r, results="asis"}
LD_values<-fread("/home/jnrunge/data/trd/LD_data/LD_plot_data.csv.gz")
cross_value<-cross
for(id in pull(TRD_regions, ID)){
    LD_values_slice<-filter(LD_values, cross==cross_value, ID==id)
    cat("### TRD region ",id)
    cat("\n") 
    cat(paste0("The LD value in the TRD region across all strains is ", pull(LD_values_slice, LD), " (R^2), which represents a percentile an SD shift of ", pull(LD_values_slice, LD_sd_multi_1), " compared to the lower-value group and a shift of ", pull(LD_values_slice, LD_sd_multi_2), " compared to the higher value group (LD is bimodally distributed)"))
     cat("\n") 
    print(readRDS(paste0("/home/jnrunge/data/trd/LD_data","/",cross,".", id,".",
                     "LD_plot",".RDS")))
     cat("\n") 
      cat("\n") 

}
````


# Positioning of genomic signals among other strains

To investigate whether the TRD regions in this strain are particularly unique among all TRD regions analyzed across strains, 

````{r, results="asis"}
cross_value<-cross
pca_df<-readRDS("~/data/trd/genomic_signals_pca.RDS")
pca_df<-mutate(pca_df, ID_thiscross=ifelse(cross==cross_value, ID, NA))
print(ggplot(pca_df,
      aes(PC1,PC2,color=cross==cross_value))+
geom_point()+geom_label(mapping=aes(label=ID_thiscross)))


````



# GO terms


````{r, results="asis"}

for(id in pull(TRD_regions, ID)){
    cat("## TRD region ",id)
    cat("\n") 
    
    if(!file.exists(paste0("/home/jnrunge/data/trd/GO_data","/",cross, ".", id, ".reducedTerms.RDS"))){
    next
    }

    GO_data<-readRDS(paste0("/home/jnrunge/data/trd/GO_data","/",cross, ".", id, ".reducedTerms.RDS"))
    
    treemapPlot(GO_data)
     cat("\n") 
    wordcloudPlot(GO_data, min.freq=1, colors="black")

 cat("\n")  
 cat("\n") 

}

````